---
title: "R Programming"
author: "Benson Nyota"
date: "2024-03-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Programming

## R Markdown presentation

We will be using R Markdown for our outline

R is optimized for vector operations and as well, it is an open source software

### CRAN :- Comprehensive R Archive Network

loading the data sets

```{r}

library(datasets)
head(iris)
```

To see the summary of the data set "iris"

```{r}
summary(iris)
```

Sometimes we can even get to visualize the data directly

```{r}
plot(iris)
```

We can install one packages that carries it many packages we would need

```{r}
install.packages("pacman")
```

```{r}
require("pacman")
```

There are two types of packages in R

-   **Base Packages** - download with r (inbuilt) but not loaded o the environment

-   **Contributed packages** - they must be downloaded

#### To load the entire collection

```{r}
pacman::p_load(pacman, dplyr, GGally, ggplot2, ggthemes, ggris, lubridate, plotly, rio, rmarkdown, shiny, stringr, tidyr)
```

### Basic Graphics

we can plot categorical data

```{r}
head(iris)
```

```{r}
plot(iris$Species)
```

-   quantitative variable

```{r}
plot(iris$Petal.Length)
```

Or else plot both quantitative data

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width)
```

categorical vs quantitative

```{r}
plot(iris$Species, iris$Petal.Width, main="Species vs Petal Width")
plot(iris$Species, iris$Petal.Length)
```

coloring the graphs

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width,
     col = "#cc0000",
     pch =19, #shape of the pointers for this case is circles
     main = "Sepa Length vs Sepal Width",
     xlab = "Sepal length fro iris dataset",
     ylab = "sepal width from iris dataset")
```

### Bar charts

we use mtcars as our datset

```{r}
head(mtcars)
```

```{r}
summary(mtcars)
```

Number of rows

```{r}
paste(nrow(mtcars), "rows")
```

Number of columns

```{r}
paste(ncol(mtcars), "Columns")
```

```{r}
str(mtcars)
```

Factoring the vs column

```{r}
mtcars$vs <- factor(mtcars$vs)
```

```{r}

#This is more of like counting distinct to create a table 
cylinders <- table(mtcars$cyl)

barplot(cylinders)
```

### Histogram

```{r}
hist(iris$Sepal.Length)
```

```{r}
hist(iris$Sepal.Width,
     main="Sepal Width Histogram")
```

Iris data

```{r}
summary(iris$Species)
```

We can arrange the graphs into different rows and columns

```{r}
par(mfrow = c(3,1)) #3 rows and 1 columns
#plot histogram for each species

hist(iris$Petal.Width [iris$Species=="setosa"], 
     xlim = c(0,3),
     breaks = 9,
     main = "Petal Width for Setosa",
     xlab = "Petal Width",
     col ="green")

hist(iris$Petal.Width [iris$Species=="versicolor"], 
     xlim = c(0,3), #xlim?
     breaks = 9,
     main = "Petal Width for Versicolor",
     xlab = "Petal Width",
     col = "red")

hist(iris$Petal.Width [iris$Species=="virginica"], 
     xlim = c(0,3),
     breaks = 9,
     main = "Petal Width for Virginica",
     xlab = "Petal Width",
     col = "blue")
```

Restoring the graphics parameters

```{r}
par(mfrow=c(1,1))
```

### Scatter plot

This case you are plotting between two quantitative variables

```{r}
plot(mtcars$wt, mtcars$mpg,
     main = "MTCARS",
     xlab ="Weights",
     ylab ="mpg")
```

adding some more plots

```{r}
plot(mtcars$wt, mtcars$mpg,
     pch =19,
     cex = 1.5, #circumference of the circle
     col = "violet",
     main = "MTCARS",
     xlab ="Weights",
     ylab ="mpg")
```

### Overlaying

```{r}
hist(lynx)
```

### Basic Statistics

These are statistical summaries of a dataset

```{r}
summary(iris$Species)
```

```{r}
summary(iris$Sepal.Length)
```

```{r}
summary(iris)
```

To be able to use describe() we need to load psych package

```{r}
p_load(psych)
```

```{r}
describe(iris$Sepal.Length)
```

From the above output, we can see that describe function gives more information about a variable.

```{r}
describe(iris)
```

### Selecting Cases

```{r}
library(datasets)

hist(iris$Petal.Length,
     main="Petal Length",
     col = "blue")
```

```{r}
summary(iris$Petal.Length)
```

##### Histogram: Selection by Category

```{r}
hist(iris$Petal.Length[iris$Species=="versicolor"],
     main="Petal Length of Versicolor Speies",
     xlab = "Petal Length")
```

Selecting by value

```{r}
hist(iris$Petal.Length[iris$Petal.Length<2],
     main = "Petal Lengths less than 2",
     xlab = "Petal Length")
```

Multiselection cases

```{r}
hist(iris$Petal.Length[iris$Species=="virginica" & iris$Petal.Length<5.5],
     main="Virginica speccies and Petal Length> 5.5",
     xlab = "Petal Length")
```

### Subset

we can create sub-dataframe by slicing out the main dataframe

```{r}
#we will be slicing the species setosa
i.setosa <- iris[iris$Species =="setosa", ]  #we put a comma to species it returns all the columns
```

```{r}
head(i.setosa)
```

### Accessing data

data type - level of measurement of a variable

they include;

-   integer

-   single & double

-   character

-   logical

-   complex

*Vector* - 1D array, all of the same data type

*Matrix* - 2D data, same length, same class

*Data frame* - Can have vectors of multiple types

*List* - most flexible, ordered collection of elements

```{r}
n1 <- 15
typeof(n1)
```

```{r}
c1 <- "c"
typeof(c1)
```

```{r}
l1 <- TRUE
typeof(l1)
```

```{r}
l2 <- F
typeof(l2)
```

### Vector

```{r}
v1 <- c(1, 2, 3, 4, 5)
is.vector(v1)
```

### Matrix

```{r}
m1 <- matrix(c(T, T, T, F, F, F), nrow=2)
is.matrix(m1)
```

### Array

```{r}
a1 <- array(c(1:24), c(4, 3, 2))

# 4 - number of rows
# 3 - number of columns
# 2 - number of tables

is.array(a1)
```

### dataframe

```{r}
n1 <- c(1, 2, 3, 4)
c1 <- c("a", "b", "c", "d")
l1 <- c(T, F, T, F)

df <- cbind(n1, c1, l1) # This returns a matrix because one has used column_bind function only to concatenate the three arrays
df1 <- as.data.frame(cbind(n1, c1, l1))
```

```{r}
df1
```

Another way of creating a data franme is directly from a matrix.

```{r}
df2 <- as.data.frame(matrix(1:9, nrow=3))
df2
```

### Factors

categories and names

lets create data

```{r}
x1 <- 1:3
y1 <- 1:9
#combine the variables
df3 <- cbind.data.frame(x1, y1)
typeof(df3$x1)
```

```{r}
str(df3)
```

Factoring

```{r}
x2 <- as.factor(c(1:3))
df4 <- as.data.frame(x2)
str(df4)
```

Restructuring the levels

```{r}
df4$x2 <- factor(df4$x2, levels = c(1, 3, 2))
str(df4)
```

Changing levels to labels

```{r}
#using df3
df4 <- df3 
df4
```

we can give labels to categorial variables

```{r}
df4$x1 <- factor(df4$x1, levels = c(1, 2, 3), labels = c("macOS", "Windows", "Linux"))
```

```{r}
typeof(df4$x1)
```

```{r}
str(df4)
```

```{r}
df4
```

Its possible to re-order

### Entering data

We can use different methods:

-   colon

-   c *for combine or concatenate*

-   seq

-   rep

*NB: \<- is the assignment operator in R*

Colon

```{r}
x1 <- 0:10
x2 <- 10:0
x1
x2
```

seq

```{r}
x3 <- seq(10)
x3
```

```{r}
x4 <- seq(30,0, by=-3) #by is for steps
x4
```

scan

```{r}
x6 <- scan() #check the console to input the figures, every entry click enter ,  to stop inputting double click enter
x6
```

rep

```{r}
x7 <- rep(TRUE, 5)
x7
```

```{r}
x8 <- rep(c(TRUE, FALSE), 5)
x8
```

```{r}
x9 <- rep(c(TRUE, FALSE), each=5)
x9
```

### Import data

rio (r input output)

```{r}
data <- import("/cloud/project/200601.csv")
```

```{r}
data
```

## Modelling Data

Hierachical clustering

technique is measuring the euclidean distance

we use inbuild data set

```{r}
#select specific columns
cars <- mtcars[, c(1:4, 6:7, 9:11)]
```

Compute and plot clusters

```{r}
hc <- cars %>% #data
  dist %>% #compute distance
  hclust
```

plotting the cluster dendrogram

```{r}
plot(hc)
```

If we would want to draw boxes

```{r}
#rect.hclust(hc, k=2, border = "gray")
#rect.hclust(hc, k=2, border = "blue")
#rect.hclust(hc, k=2, border = "red")
#rect.hclust(hc, k=2, border = "green")
```

### Principal Component

Dimensionality Reduction

PCA

This is the aspect of reducing dimensionality and maintaining the meaning.

```{r}
#using cars subset
# we can compute PC
pc <- prcomp(cars, 
             center = TRUE, 
             scale = TRUE)
```

```{r}
summary(pc)
```

```{r}
plot(pc)
```

```{r}
predict(pc) %>%
  round(2)
```
```{r}
biplot(pc)
```

### Regression

```{r}
# we have already loaded our pacman
#Load dataset
library(datasets)
data <-  USJudgeRatings

#define variable groups
x <- as.matrix(data[-12])
y <- data[, 12]
```

```{r}
data
```

Regression with simultaneous entry

```{r}
reg <- lm(y~x)
reg
```
```{r}
summary(reg)
```
```{r}
hist(residuals(reg))
```
More Models
```{r}
p_load(lars, caret)
# lars - least angle regression
# caret - classification and regression training

stepwise <- lars(x,y, type="stepwise")
stepwise
```



```{r}
forward <- lars(x,y, type="forward.stagewise")
forward
```
```{r}
lasso <- lars(x,y, type="lasso")
lasso
```

```{r}
lar <- lars(x,y, type="lar")
lar
```

```{r}
r2comp <- c(stepwise$R2[6], forward$R2[6],
            lar$R2[6], lasso$R2[6] %>% round(2))

r2comp
```

```{r}
names(r2comp) <- c("stepwise", "forward", "lar", "lasso")
```


### Converting data type

```{r}
x <- 400
y <- "100"

print(x + strtoi(y))
```

