---
title: "R Programming"
author: "Benson Nyota"
date: "2024-03-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Programming

## R Markdown presentation

We will be using R Markdown for our outline

R is optimized for vector operations and as well, it is an open source software

### CRAN :- Comprehensive R Archive Network

loading the data sets

```{r}

library(datasets)
head(iris)
```

To see the summary of the data set "iris"

```{r}
summary(iris)
```

Sometimes we can even get to visualize the data directly

```{r}
plot(iris)
```

We can install one packages that carries it many packages we would need

```{r}
install.packages("pacman")
```

```{r}
require("pacman")
```

There are two types of packages in R

-   **Base Packages** - download with r (inbuilt) but not loaded o the environment

-   **Contributed packages** - they must be downloaded

#### To load the entire collection

```{r}
pacman::p_load(pacman, dplyr, GGally, ggplot2, ggthemes, ggris, lubridate, plotly, rio, rmarkdown, shiny, stringr, tidyr)
```

### Basic Graphics

we can plot categorical data

```{r}
head(iris)
```

```{r}
plot(iris$Species)
```

-   quantitative variable

```{r}
plot(iris$Petal.Length)
```

Or else plot both quantitative data

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width)
```

categorical vs quantitative

```{r}
plot(iris$Species, iris$Petal.Width, main="Species vs Petal Width")
plot(iris$Species, iris$Petal.Length)
```

coloring the graphs

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width,
     col = "#cc0000",
     pch =19, #shape of the pointers for this case is circles
     main = "Sepa Length vs Sepal Width",
     xlab = "Sepal length fro iris dataset",
     ylab = "sepal width from iris dataset")
```

### Bar charts

we use mtcars as our datset

```{r}
head(mtcars)
```

```{r}
summary(mtcars)
```

Number of rows

```{r}
paste(nrow(mtcars), "rows")
```

Number of columns

```{r}
paste(ncol(mtcars), "Columns")
```

```{r}
str(mtcars)
```

Factoring the vs column

```{r}
mtcars$vs <- factor(mtcars$vs)
```

```{r}

#This is more of like counting distinct to create a table 
cylinders <- table(mtcars$cyl)

barplot(cylinders)
```

### Histogram

```{r}
hist(iris$Sepal.Length)
```

```{r}
hist(iris$Sepal.Width,
     main="Sepal Width Histogram")
```

Iris data

```{r}
summary(iris$Species)
```

We can arrange the graphs into different rows and columns

```{r}
par(mfrow = c(3,1)) #3 rows and 1 columns
#plot histogram for each species

hist(iris$Petal.Width [iris$Species=="setosa"], 
     xlim = c(0,3),
     breaks = 9,
     main = "Petal Width for Setosa",
     xlab = "Petal Width",
     col ="green")

hist(iris$Petal.Width [iris$Species=="versicolor"], 
     xlim = c(0,3), #xlim?
     breaks = 9,
     main = "Petal Width for Versicolor",
     xlab = "Petal Width",
     col = "red")

hist(iris$Petal.Width [iris$Species=="virginica"], 
     xlim = c(0,3),
     breaks = 9,
     main = "Petal Width for Virginica",
     xlab = "Petal Width",
     col = "blue")
```

Restoring the graphics parameters

```{r}
par(mfrow=c(1,1))
```

### Scatter plot

This case you are plotting between two quantitative variables

```{r}
plot(mtcars$wt, mtcars$mpg,
     main = "MTCARS",
     xlab ="Weights",
     ylab ="mpg")
```

adding some more plots

```{r}
plot(mtcars$wt, mtcars$mpg,
     pch =19,
     cex = 1.5, #circumference of the circle
     col = "violet",
     main = "MTCARS",
     xlab ="Weights",
     ylab ="mpg")
```
